Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-05-29 conclusions.tex

@@

In this paper, we have presented the algorithms underlying \enzo, an
open-source adaptive mesh refinement code designed for
self-gravitating compressible fluid dynamics, including the effects of
magnetic fields, radiation transport, and a variety of microphysical
and subgrid processes.  In addition, we have described the \enzo\ code
development process, have shown the outputs of a set of representative
test problems, and have provided information about \enzo's performance
and parallel scaling on recent supercomputing platforms.  The \enzo\
code, its test suite, and all of the scripts used to generate plots
and figures for this paper are open source and are available at the
\enzo\ website, \url{http://enzo-project.org}.  Furthermore, the yt toolkit,
which is designed to analyze \enzo\ data (as well as data from a wide
variety of other simulation tools), can be found at its website,
\url{http://yt-project.org}.  Both of these codes have active user and
developer communities, extensive documentation and user suppport, and
strong mechanisms for users to contribute their changes and fixes to
the codebase.

@@

In the longer term, it seems likely that trends in supercomputing
hardware will require substantial modifications to \enzo's core
infrastructure, and very possibly some of the core algorithsms.  More
specifically, these trends involve the usage of specialized
large-core-count, vectorized computing units such as graphics
processuing units (GPUs) or chips like the Intel Xeon Phi, as well as
precipitously decreasing amounts of RAM per computing core.  The
former trend means that the amount of processing power per compute
node will continue to trend upward, likely much faster than the
bandwidth between nodes, and will require tremendous reduction in
inter-node (and possibly inter-CPU) communication in order to maintain
code scalability.  Also, much of the current code will need to be
rewritten to take advantage of the vector nature of these CPUs, making
assumptions that are quite unlike those made in much of the current
codebase.  The latter trend means that duplication of data -- for
example, the grid hierarchy -- must be effectively eliminated to save
memory, and all inter-core and inter-node communication must be
carefully thought through to minimize the amount of data moved.  An
additional challenge as one goes to core counts in the tens to
hundreds of millions (or more) is that the reliability of individual
computing elements will become much more of an issue, requiring
robustness to hardware failure to be built into the code at some
level.  Furthermore, we are nearing the physical limits of transistor
speed and latency~\citep{feynman1999feynman}, meaning that simple
hardware improvements will not make these challenges disappear, and
careful thought (and the rewriting of a great deal of code) must take
place!  These challenges are not unique to the \enzo\ code, and in
fact are faced by effectively all applications that wish to take
advantage of new computational architectures.  Thus, we suspect that
\enzo\ (or a code that has the capabilities of \enzo, from a user's
point of view) will continue to be usable at the largest scales on
such machines.

>>>>

% EJT added intro sentense to section on future developement and tweaked final sentense

@@

In this paper, we have presented the algorithms underlying \enzo: an
open-source adaptive mesh refinement code designed for
self-gravitating compressible fluid dynamics, including the effects of
magnetic fields, radiation transport, and a variety of microphysical
and subgrid processes.  In addition, we have described the \enzo\ code
development process, have shown the outputs of a set of representative
test problems, and have provided information about \enzo's performance
and parallel scaling on recent supercomputing platforms.  The \enzo\
code, its test suite, and all of the scripts used to generate plots
and figures for this paper are open source and are available at the
\enzo\ website, \url{http://enzo-project.org}.  Furthermore, the \texttt{yt} toolkit,
which is designed to analyze \enzo\ data (as well as data from a wide
variety of other simulation tools), can be found at its website,
\url{http://yt-project.org}.  Both of these codes have active user and
developer communities, extensive documentation and user support, and
strong mechanisms for users to contribute their changes and fixes to
the codebase.

@@

With the continual rapid development of computer hardware, it makes sense to review
not only at \enzo's current capabilities, but to look towards its future development 
in view of predicted technological trends. These trends in supercomputing hardware suggest that  
substantial modifications to \enzo's core infrastructure, and very possibly some of 
the core algorithms will be required. More
specifically, the progression involves the usage of specialized
large-core-count, vectorized computing units such as graphics
processing units (GPUs) or chips like the Intel Xeon Phi, as well as
precipitously decreasing amounts of RAM per computing core.  The
former trend means that the amount of processing power per compute
node will continue to increase, likely much faster than the
bandwidth between nodes, and will require tremendous reduction in
inter-node (and possibly inter-CPU) communication in order to maintain
code scalability.  Also, much of the current code will need to be
rewritten to take advantage of the vector nature of these CPUs, making
assumptions that are quite unlike those made in much of the current
codebase.  The latter trend means that duplication of data -- for
example, the grid hierarchy -- must be effectively eliminated to save
memory, and all inter-core and inter-node communication must be
carefully thought through to minimize the amount of data moved.  An
additional challenge as one goes to core counts in the tens to
hundreds of millions (or more) is that the reliability of individual
computing elements will become much more of an issue, requiring
robustness to hardware failure to be built into the code at some
level.  Furthermore, we are nearing the physical limits of transistor
speed and latency~\citep{feynman1999feynman}, meaning that simple
hardware improvements will not make these challenges disappear, and
careful thought (and the rewriting of a great deal of code) must take
place!  These challenges are not unique to the \enzo\ code, and in
fact are faced by effectively all applications that wish to take
advantage of new computational architectures. We therefore anticipated that 
\enzo\ (or a code that has the capabilities of \enzo, from a user's
point of view) will continue to develop to be usable at the largest scales on
such machines.

### 2013-05-29 parallel.tex

Many astrophysical problems, such as cosmological galaxy formation
\citep{2012ApJ...749..140H}, high-resolution disk galaxy simulations
\citep{}, high-redshift \citep{2009Sci...325..601T}, and present-day
star formation \citep{Collins12a}, involve multi-scale physics that
span several orders of magnitude in both space and time.  In these
situations, using Enzo in its adaptive mesh refinement mode is
beneficial.  Because the adaptive grid hierarchy is dynamic, grid
boundaries and, thus, communication patterns can be unpredictable,
hindering strong scaling behavior to high core counts.

>>>>

Many astrophysical problems, such as cosmological galaxy formation
\citep{2012ApJ...749..140H}, high-resolution disk galaxy simulations
\citep{2011ApJ...738...54K}, high-redshift \citep{2009Sci...325..601T}, and present-day
star formation \citep{Collins12a}, involve multi-scale physics that
span several orders of magnitude in both space and time.  In these
situations, using Enzo in its adaptive mesh refinement mode is
beneficial.  Because the adaptive grid hierarchy is dynamic, grid
boundaries and, thus, communication patterns can be unpredictable,
hindering strong scaling behavior to high core counts.

### 2013-05-29 tests-orszagtang.tex

\subsubsection{MHD: Orszag-Tang Vortex}
\label{sec.tests.mhd}
Figure~\ref{fig.orszag} shows the Orszag-Tang vortex problem, a
standard magnetohydrodynamic test problem \citep{Orszag79}.  The left panel shows
the result using \enzo's constrained transport MHD method, while the right panel shows the result using
\enzo's implementation of Dedner MHD.
The Orszag-Tang vortex test shows that significant small scale structure can be generated in MHD
from large scale initial perturbations, and is used to compare the
effective resolution of different MHD schemes.  The test begins with uniform
density, $\rho_0=25/36 \pi$ and pressure, $P_0=5/12 \pi$ (as with
other tests in this section, in the absence of gravity or chemistry we use
arbitrary units).  There is a
single rotational mode in the velocity, and two in the magnetic field:
${\bf v}_0 $ = (-sin(2$\pi$ y) $ \hat{x}$ , sin(2$\pi$ x) $\hat{y}$),
${\bf B}_0$ = (-sin(2 $\pi$ y) $ \hat{x},$ sin( 4 $\pi$ x )$\hat{y}$).
The simulation is evolved to $t=0.48$ (in arbitrary units).  One can see that the structures
are accurately represented as compared to, for instance,
\citet{Toth00}, and that the resolution of shocks is comparable in
both methods.

>>>>

\subsubsection{MHD: Orszag-Tang Vortex}
\label{sec.tests.mhd}
Figure~\ref{fig.orszag} shows the Orszag-Tang vortex problem, a
classic MHD test problem described by \citep{Orszag79}.  The left panel shows
the result using \enzo's constrained transport MHD method, while the right panel shows the result using
\enzo's implementation of Dedner MHD.
The Orszag-Tang vortex test shows that significant small scale structure can be generated in MHD
from large scale initial perturbations, and is used to compare the
effective resolution of different MHD schemes.  The test begins with uniform
density, $\rho_0=25/36 \pi$ and pressure, $P_0=5/12 \pi$ (as with
other tests in this section, in the absence of gravity or chemistry we use
arbitrary units).  There is a
single rotational mode in the velocity, and two in the magnetic field:
${\bf v}_0 $ = (-sin(2$\pi$ y) $ \hat{x}$ , sin(2$\pi$ x) $\hat{y}$),
${\bf B}_0$ = (-sin(2 $\pi$ y) $ \hat{x},$ sin( 4 $\pi$ x )$\hat{y}$).
The simulation is evolved to $t=0.48$ (in arbitrary units).  One can see that the structures
are accurately represented as compared to, for example,
\citet{Toth00}, and that the resolution of shocks is comparable in
both methods.

### 2013-05-29 tests-pointsourcegravity.tex

This test is a simple test of the ability of the Poisson solver to correctly reproduce the acceleration around a point source without any dynamics.  We place a single particle in the center of the unit domain with root grid dimensions $32^3$ and add a static AMR grid with refinement factor two that covers the central $1/8^3$ of the domain (i.e. so the refined grid is only $8^3$).  We then place 5000 particles throughout the domain distributed uniformly in angle and with the logarithm of the radius (from the central point).  The measured accelerations are shown in Figure~\ref{fig.gravitytest}.  As is typical for Particle-Mesh based methods, the errors decrease at large distances and peak on roughly the grid scale.  The accelerations are relatively smooth across the top grid/fine grid boundary with the most noticeable impact being a drop in the amplitude of tangential errors on the fine grid.  The force law rolls over at about 1.6 fine cell widths (equivalent to 0.8 root grid widths in this figure), consistent with the CIC-deposition and interpolation used.

>>>>

This is a simple test of the ability of the Poisson solver to
correctly reproduce the acceleration around a point source without any
dynamics.  We place a single particle in the center of the unit domain
with root grid dimensions $32^3$ and add a static, nested grid 
refinemed by a factor of two that covers the central $1/8^3$ of the
domain (i.e. so the refined grid is only $8^3$).  We then place 5000
particles throughout the domain distributed uniformly in angle and
in the logarithm of the radius (from the central point).  The
measured accelerations are shown in Figure~\ref{fig.gravitytest}.  As
is typical for Particle-Mesh based methods, the errors decrease at
large distances and peak on roughly the grid scale.  The accelerations
are relatively smooth across the top grid/fine grid boundary with the
most noticeable impact being a drop in the amplitude of tangential
errors on the fine grid.  The force law rolls over at about 1.6 fine
cell widths (equivalent to 0.8 root grid widths in this figure),
consistent with the CIC-deposition and interpolation used.

### 2013-05-29 tests-sedov.tex

\begin{figure}
\begin{center}
\includegraphics[width=0.4\textwidth]{figures/sedov-ppm-slice.eps}
\includegraphics[width=0.4\textwidth]{figures/sedov-zeus-slice.eps}
\caption{Density slice from the Sedov Blast test at $t = 0.07$. Left:
results using the Piecewise Parabolic Method hydro scheme.  Right: results using the \zeus\
hydro method. Notably, the \zeus\ shock front has progressed less far
than in the PPM run. This is due to energy loss when conserving only
internal, and not total, energy.}
\label{fig.sedov1}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{figures/sedov-profiles.eps}
\caption{Radially averaged profiles for the Sedov Blast test at $t =
0.07$. Clockwise from top-left shows density, velocity, internal
energy and pressure.  The black solid line shows the analytic
solution.  The blue dashed line shows the simulation using the PPM
method, and the red dot-dashed line using \zeus.  The \zeus\ result
substantially lags the true result due to total energy not being
explicitly converged.}
\label{fig.sedov2}
\end{center}
\end{figure}

The Sedov Blast Test \citep{Sedov1959} models an intense explosion,
initiated by depositing thermal energy into a homogenous distribution
of gas. The result is a strong spherical shock wave centered on the
point of energy injection.  This problem is a popular test of
astrophysical numerical codes for three reasons: First, it is
particularly appropriate to astrophysics since it represents the
situation when a supernovae explosion occurs. Second, it has an
exact analytical solution whereby the shock front's radial position is
given by:

@@

\noindent where $E_0$ is the initial energy injection, $\rho_0$ is the
background density and $\alpha = 1.0$ for cylindrical symmetry and an
ideal gas with $\gamma = 1.4$. For the full derivation see
\citet{Sedov1959}. This solution makes it possible to assess how well
the code performs.  Third, the spherical shock is a challenging
problem for numerical codes (both grid and particle) since the shock
wave increases in size as the simulation progresses and its
symmetrical nature highlights any directional preferences that grid
codes can succumb to. The test presented here is the two-dimensional
version that is included in the \enzo\ distribution. The
three-dimensional results from this test, both for \enzo\ and three
other leading astrophysics codes, can be founds in \citet{Tasker2008}.

In the initial state, the box contains a homogenous distribution of
gas at a density of 1 (note that, in the absense of gravity, this
problem is scale-free and thus without units). Thermal energy is
deposited into a single cell at the center of the box with $E_0 =
10.0$. The problem is completed in two dimensions with reflecting
boundary conditions (the default for \enzo) and uses a box size of side
$1$. For this problem, we selected a top grid of $100 \times 100$
cells and a maximum of four levels of refinement, placed based on
shock location and the slope in density and total energy. The
exception to this scheme is in the initial conditions, where grids are
placed directly around the injection point. The results were assessed
at $t = 0.07$, which corresponding to a time just before the shock
reaches the box edge (see Figure~\ref{fig.sedov1}).

Figure~\ref{fig.sedov2} shows the radial profiles for the simulation
run with the PPM hydro-solver (blue dashed line) and the \zeus\
hydro-solver (red dash-dot-dot line) together with the analytical
solution (black solid line). Clockwise from the top-left are density,
velocity, internal energy and pressure. PPM matches the analytical
solution extremely well for all quantities. However, the shock front
in the \zeus\ simulation lags behind the analytical position. This can
also be seen in the slices shown in Figure~\ref{fig.sedov1}. The cause
of this discrepancy is that \zeus\ shows a substantial energy loss
during the first few timesteps and produces a diamond-shaped, rather
than spherical, shockfront during this time. After this, the code
correctly conserves energy but this intial energy loss remains clearly
visible in the position of the shock at $t = 0.07$. This problem was
addressed directly by \citet{Clarke2010}, who attributed the source of
the issue to this version of \zeus\ solving the internal, rather than
total, energy equation. In situations with strong energy gradients,
this choice caused an energy loss and the artificial viscosity
produces the direction-dependent shockfront shape. In their paper,
\citet{Clarke2010} presents results from an alternative version of
\zeus\ that conserves total energy. This problem is less marked for
smaller energy gradients and it should be noted that the \zeus\ hydro
algorithm's stability and speed make it a highly competitive choice,
despite the disagreements in this test.

>>>>

\begin{figure}
\begin{center}
\includegraphics[width=0.4\textwidth]{figures/sedov-ppm-slice.eps}
\includegraphics[width=0.4\textwidth]{figures/sedov-zeus-slice.eps}
\caption{Density slices from the Sedov Blast test at $t = 0.07$. Left:
results using the Piecewise Parabolic Method hydro scheme.  Right: results using the \zeus\
hydro method. Notably, the \zeus\ shock front has progressed less far
than in the PPM run. This is due to energy loss when conserving only
internal, and not total, energy.}
\label{fig.sedov1}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{figures/sedov-profiles.eps}
\caption{Radial profiles for the Sedov Blast test at $t =
0.07$. Clockwise from top-left shows density, velocity, internal
energy and pressure.  The black solid line shows the analytic
solution.  The blue dashed line shows the simulation using the PPM
method, and the red dot-dashed line using \zeus.  The \zeus\ result
substantially lags the true result due to total energy not being
explicitly converged.}
\label{fig.sedov2}
\end{center}
\end{figure}

The Sedov Blast Test \citep{Sedov1959} models an intense explosion,
initiated by depositing thermal energy into a homogenous distribution
of gas. The result is a strong spherical shock wave centered on the
point of energy injection.  This problem is a popular test of
astrophysical simulation codes for three reasons.  First, this problem
is representative of the astrophysical phenomenon of a supernova
explosion.   Second, an analytical solution to the evolution of the
Sedov blast-wave exists, allowing for a direct test of the accuracy of
the code.  The radius of the shock front as a function of time is
given by

@@

\noindent where $E_0$ is the initial energy injection, $\rho_0$ is the
background density and $\alpha = 1.0$ for cylindrical symmetry and an
ideal gas with $\gamma = 1.4$.  For the full derivation see
\citet{Sedov1959}.  Third, as the spherical shock expands, its
symmetry, or lack thereof, serves to highlight any directional
preferences of the hydrodynamics solvers.  The test presented here is
the two-dimensional 
version that is included in the \enzo\ distribution. The
three-dimensional results from this test, both for \enzo\ and three
other leading astrophysics codes, can be found in \citet{Tasker2008}.

In the initial state, the box contains a homogenous distribution of
gas at a density of 1 (note that, in the absense of gravity, this
problem is scale-free and thus without units). Thermal energy is
deposited into a single cell at the center of the box with $E_0 =
10.0$. The problem is run in two dimensions with reflecting
boundary conditions and a box size of side
$1$.  We selecte a top grid of $100 \times 100$
cells and a maximum of four levels of adaptive refinement, refining on
shocks and the slopes of the density and total energy fields. The
exception to this scheme is in the initial conditions, where
additional refinement is placed directly around the injection point to
better resolve spherical geometry. The results are assessed
at $t = 0.07$, which corresponds to a time just before the shock
reaches the box edge (see Figure~\ref{fig.sedov1}).

Figure~\ref{fig.sedov2} shows the radial profiles for the simulation
run with the PPM hydro-solver (blue dashed line) and the \zeus\
hydro-solver (red dash-dot-dot line) together with the analytical
solution (black solid line). Clockwise from the top-left are density,
velocity, internal energy and pressure. PPM matches the analytical
solution extremely well for all quantities. However, the shock front
in the \zeus\ simulation lags behind the analytical position. This can
also be seen in the slices shown in Figure~\ref{fig.sedov1}. The cause
of this discrepancy is that \zeus\ shows a substantial energy loss
during the first few timesteps and produces a diamond-shaped, rather
than spherical, shockfront during this time. After this, the code
correctly conserves energy but this intial energy loss remains clearly
visible in the position of the shock at $t = 0.07$. This problem was
addressed directly by \citet{Clarke2010}, who attributed the source of
the issue to this version of \zeus\ solving the internal, rather than
total, energy equation. In situations with strong energy gradients,
this choice caused an energy loss and the artificial viscosity
produces the direction-dependent shockfront shape. In their paper,
\citet{Clarke2010} present results from an alternative version of
\zeus\ that conserves total energy. This problem is less marked for
smaller energy gradients and it should be noted that the \zeus\ hydro
algorithm's stability and speed make it a highly competitive choice,
despite the disagreements in this test.
